---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# AdaptiveDesignOptimizerSparseLP

<!-- badges: start -->
[![Travis build status](https://travis-ci.com/mrosenblum/AdaptiveDesignOptimizerSparseLP.svg?branch=master)](https://travis-ci.com/mrosenblum/AdaptiveDesignOptimizerSparseLP)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/mrosenblum/AdaptiveDesignOptimizerSparseLP?branch=master&svg=true)](https://ci.appveyor.com/project/mrosenblum/AdaptiveDesignOptimizerSparseLP)
<!-- badges: end -->

The goal of AdaptiveDesignOptimizerSparseLP is to construct an optimal two-Stage, two subpopulation, adaptive enrichment designs for a given problem. The problem inputs are the desired familywise Type I error rate and power, and set of scenarios (data generating distributions) of interest. The software aims to minimize the expected sample size under the power and Type I error constraints.

## Installation

You can install the released version of AdaptiveDesignOptimizerSparseLP from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("AdaptiveDesignOptimizerSparseLP")
```

## Example

This is an example, which involves solving the problem from Example 3.2 as described in Section 5.2 of the manuscript. 

```{r example}
#Install R package if not already done so:
#remotes::install_github("mrosenblum/AdaptiveDesignOptimizerSparseLP")

# Load R package
library(AdaptiveDesignOptimizerSparseLP)

#This problem was solved in 5 iterations, with iteration based on refining the decision/rejection regions partition based on the solution from the previous iteration. Each iteration involves specifying and then solving the corresponding linear program. 

# Set all problem input parameters based on Example 3.2, but using explicit choices of sample sizes defined by setting n=200:
subpopulation.1.proportion=0.5;
total.alpha=0.05-(1e-4);
data.generating.distributions=matrix(data=c(0,0,1,1,1,1,
                                           0,sqrt(1/2)*(qnorm(0.95+1e-4)+qnorm(0.95))/5,1,1,1,1,
                                           sqrt(1/2)*(qnorm(0.95+1e-4)+qnorm(0.95))/5,0,1,1,1,1,
                                           sqrt(1/2)*(qnorm(0.95+1e-4)+qnorm(0.95))/5,sqrt(1/2)*(qnorm(0.95+1e-4)+qnorm(0.95))/5,1,1,1,1),nrow=4,ncol=6,byrow=TRUE,dimnames=list(c(),c("Delta1","Delta2","Variance10","Variance11","Variance20","Variance21")));
stage.1.sample.sizes=c(50,50);
stage.2.sample.sizes.per.enrollment.choice=matrix(c(50,50,
                                                   0,0,
                                                   150,0,
                                                   0,150),nrow=4,ncol=2,byrow=TRUE,dimnames=list(c(),c("Subpopulation1Stage2SampleSize","Subpopulation2Stage2SampleSize")));
objective.function.weights=0.25*c(1,1,1,1);
prior.covariance.matrix=diag(2);
power.constraints=matrix(c(0,0,0,
			   0,0.83,0,
			   0.83,0,0,
			   0,0,0.83),nrow=4,ncol=3,byrow=TRUE,dimnames=list(c(),c("PowerH01","PowerH02","PowerH0C")));
type.of.LP.solver="matlab";
discretization.parameter=c(1,1,10);
number.cores=30;

# Run first iteration solving sparse linear program at slightly stronger power constraint (0.83) than the intended final output (0.82)
###optimize_design(subpopulation.1.proportion,total.alpha,data.generating.distributions,stage.1.sample.sizes,stage.2.sample.sizes.per.enrollment.choice,objective.function.weights,power.constraints,type.of.LP.solver,discretization.parameter,number.cores,ncp.list=c(),list.of.rectangles.dec=c(),LP.iteration=1,prior.covariance.matrix)
# Construct new list of familywise Type I error constraints focusing on neighborhoods of active constraints from previous iteration 
# The Active Type I error constraints are contained in the output file optimized.design1.rdata and stored in the variable ncp.active.FWER.constraints, and in this case are the following 3 pairs: 0 0; 0.000000 1.960894; 1.960894 0.000000
ncp.list <- list(c(0,0))
for(z in seq(1.5,3,length=30)) {ncp.list <- c(ncp.list,list(c(0,z)),list(c(z,0)))}
for(z in seq(-9,9,length=30)) {ncp.list <- c(ncp.list,list(c(z,-z)))}

# Run second iteration solving sparse linear program as in first iteration but using Type I error constraints defined above
###optimize_design(subpopulation.1.proportion,total.alpha,data.generating.distributions,stage.1.sample.sizes,stage.2.sample.sizes.per.enrollment.choice,objective.function.weights,power.constraints,type.of.LP.solver,discretization.parameter,number.cores,ncp.list,list.of.rectangles.dec=c(),LP.iteration=2,prior.covariance.matrix)
# Refine Decision Region Rectangle Set
load("optimized.design2.rdata")
list.of.rectangles.dec <- refine_decision_rectangles(subpopulation.1.proportion=0.5,stage.1.sample.sizes=stage.1.sample.sizes,stage.2.sample.sizes.per.enrollment.choice=stage.2.sample.sizes.per.enrollment.choice,list.of.rectangles.dec=list.of.rectangles.dec,sln=sln,set.rectangles.with.identically.valued.neighbors.and.split.others=TRUE,LP.iteration=2)
# Construct new list of familywise Type I error constraints focusing on neighborhoods of active constraints from previous iteration 
# The Active Type I error constraints are contained in the output file optimized.design2.rdata and stored in the variable ncp.active.FWER.constraints
ncp.list <- list(c(0,0))
for(z in seq(1.5,3,length=30)) {ncp.list <- c(ncp.list,list(c(0,z)),list(c(z,0)))}
for(z in seq(-9,9,length=10)) {ncp.list <- c(ncp.list,list(c(z,-z)))}

# Run third iteration solving sparse linear program as in first iteration but using Type I error constraints defined above
###optimize_design(subpopulation.1.proportion,total.alpha,data.generating.distributions,stage.1.sample.sizes,stage.2.sample.sizes.per.enrollment.choice,objective.function.weights,power.constraints,type.of.LP.solver,discretization.parameter,number.cores,ncp.list,list.of.rectangles.dec,LP.iteration=3,prior.covariance.matrix)
# Refine Decision Region Rectangle Set
load("optimized.design3.rdata")
list.of.rectangles.dec <- refine_decision_rectangles(subpopulation.1.proportion=0.5,stage.1.sample.sizes=stage.1.sample.sizes,stage.2.sample.sizes.per.enrollment.choice=stage.2.sample.sizes.per.enrollment.choice,list.of.rectangles.dec=list.of.rectangles.dec,sln=sln,set.rectangles.with.identically.valued.neighbors.and.split.others=TRUE,LP.iteration=3)
# Construct new list of familywise Type I error constraints focusing on neighborhoods of active constraints from previous iteration 
# The Active Type I error constraints are contained in the output file optimized.design3.rdata and stored in the variable ncp.active.FWER.constraints
ncp.list <- list(c(0,0))
for(z in seq(1.8,2.5,length=20)) {ncp.list <- c(ncp.list,list(c(0,z)),list(c(z,0)))}

# Run fourth iteration solving sparse linear program as in first iteration but using Type I error constraints defined above
###optimize_design(subpopulation.1.proportion,total.alpha,data.generating.distributions,stage.1.sample.sizes,stage.2.sample.sizes.per.enrollment.choice,objective.function.weights,power.constraints,type.of.LP.solver,discretization.parameter,number.cores,ncp.list,list.of.rectangles.dec,LP.iteration=4,prior.covariance.matrix)
# Set Decision Region Rectangles 
load("optimized.design4.rdata")
list.of.rectangles.dec <- refine_decision_rectangles(subpopulation.1.proportion=0.5,stage.1.sample.sizes=stage.1.sample.sizes,stage.2.sample.sizes.per.enrollment.choice=stage.2.sample.sizes.per.enrollment.choice,list.of.rectangles.dec=list.of.rectangles.dec,sln=sln,set.rectangles.with.identically.valued.neighbors.and.split.others=FALSE,round.each.decision.rectangle.to.integer=TRUE,LP.iteration=4)
# Construct new list of familywise Type I error constraints focusing on neighborhoods of active constraints from previous iteration 
# The Active Type I error constraints are contained in the output file optimized.design4.rdata and stored in the variable ncp.active.FWER.constraints
ncp.list <- list(c(0,0))
for(z in seq(2,2.5,length=500)) {ncp.list <- c(ncp.list,list(c(0,z)),list(c(z,0)))}
for(z in seq(-9,9,length=950)) {ncp.list <- c(ncp.list,list(c(0,z)),list(c(z,0)),list(c(z,-z)))}
for(z in c(seq(-3,-1,length=500),seq(1,3,length=500))) {ncp.list <- c(ncp.list,list(c(z,-z)))}

# Run fifth iteration solving sparse linear program as in first iteration but using Type I error constraints defined above
optimize_multiple_testing_procedure(subpopulation.1.proportion,total.alpha=0.049,data.generating.distributions,stage.1.sample.sizes,stage.2.sample.sizes.per.enrollment.choice,objective.function.weights,power.constraints,type.of.LP.solver,discretization.parameter=c(1,0.25,10),number.cores,ncp.list,list.of.rectangles.dec,LP.iteration=5,prior.covariance.matrix,round.each.multiple.testing.procedure.rectangle.to.integer=TRUE)
```
